Quasi-universal acceptance of basic science in the US

### 2025-03-11 paper.Rmd
before final tweaks for re-submission

---
title             : "Quasi-universal acceptance of basic science in the US"
shorttitle        : "Does anyone mistrust basic science?"

header-includes:  | # to prevent floats from moving past certain points (for the appendix)
  \usepackage{placeins}

author: 
  - name          : ""
    affiliation   : ""

affiliation:
  - id            : ""
    institution   : ""

abstract: |
  Substantial minorities of the population report a low degree of trust in science, or endorse conspiracy theories that violate basic scientific knowledge. This might indicate a wholesale rejection of science. In four studies, we asked 782 US participants questions about trust in science, conspiracy beliefs, and basic science (e.g. the relative size of electrons and atoms). Participants were provided with the scientifically consensual answer to the basic science questions, and asked whether they accept it. Acceptance of the scientific consensus was very high in the sample as a whole (95.1%), but also in every sub-sample (e.g., no trust in science: 87.3%; complete endorsement of flat Earth theory: 87.2%). This quasi-universal acceptance of basic science suggests that people are motivated to reject specific scientific beliefs, and not science as a whole. This could be leveraged in science communication.

keywords          : Trust in science; conspiracy theories; conspiracy thinking; science knowledge.

floatsintext      : yes
linenumbers       : no 
draft             : no
mask              : no

figurelist        : no
tablelist         : no
footnotelist      : no

classoption       : "doc" # "doc" for nice look, "man" for manuscripty
output            : papaja::apa6_doc # "_doc" for word; however, note that some of the features of kableExtra are not available for word and will yield errors. For example to knit to word, you'll have to comment out all "add_header_above()" functions for model output tables, or e.g. set always allow html as true in the yaml heading

always_allow_html: true

appendix:
  - "appendix_exp1.Rmd"
  - "appendix_exp2.Rmd"
  - "appendix_exp3.Rmd"
  - "appendix_exp4.Rmd"
  - "appendix_acceptance_knowledge.Rmd"
  - "appendix_conspiracy_thinking.Rmd"
  - "appendix_trust.Rmd"
  - "appendix_conspiracy_belief.Rmd"
  - "appendix_knowledge.Rmd"
  - "appendix_by-question.Rmd"

bibliography: references.bib
---

@@

Unfortunately, people who report a high degree of trust in science are a minority in most countries, and they are outnumbered by people who have low trust in science in many areas, e.g., most of Africa and significant parts of Asia [@wellcomeglobalmonitorWellcomeGlobalMonitor2018; @wellcomeglobalmonitorWellcomeGlobalMonitor2020]. Moreover, trust in science has recently been declining in some countries [@alganTrustScientistsTimes2021; although see @wellcomeglobalmonitorPublicTrustScientists2021], including the US [@brianAmericansTrustScientists2023], where it is also increasingly polarizing [@gauchatPoliticizationSciencePublic2012; @krauseTrendsAmericansTrust2019a; @liPolarizationPublicTrust2022]. Rejection of science can take the more extreme form of conspiracy theories, many of which question the scientific consensus on vaccination, climate change, and even the shape of the Earth.

What does this apparent rejection of science actually entail? Do people who say they do not trust science, or who believe in conspiracy theories at odds with well-established science, reject most of science? Or, on the contrary, do they object to a few specific facets of science, while still accepting the overwhelming majority of basic science? Answering this question has theoretical implications: if some people reject science wholesale, then it’s possible that their distrust of science causes their rejection of specific scientific theories, or facilitates their acceptance of conspiracy theories; by contrast, if everyone trusts most of science, then the mistrust of specific facets of science is more likely to be post-hoc, a rationalization of other beliefs or behaviors. The present question also has practical implications: many communication attempts leverage the scientific consensus [e.g. on vaccination, climate change, etc., for review, see @vanstekelenburgScientificConsensusCommunicationContested2022; see also @veckalov27countryTestCommunicating2024]. These attempts are more likely to be successful if everyone trusts basic science than if some people reject science wholesale.

Studies of science knowledge provide relevant evidence, as people who answer science knowledge questions correctly presumably trust scientists in this respect. The average level of science knowledge is quite low, even in countries with extensive science education. For example, only 55% in the US (in 2014) and only 46% in the EU (in 2005) knew that antibiotics kill only bacteria and not viruses [@nationalacademiesofsciencesengineeringandmedicineScienceLiteracyConcepts2016]. However, these results only provide a lower bound on how much people trust basic science, as participants who do not provide the correct answer might accept it when it is pointed out to them.

@@

In a series of four pre-registered online studies (total n = 782), we asked US participants questions about well-established scientific facts. For each question, we asked participants what they thought the correct answer was (testing their knowledge of science), we informed them of the scientifically accepted answer, and asked them whether they accepted it (measuring their trust in basic science). We also measured participants’ trust in science using standard measures, as well as their beliefs in various conspiracy theories and their tendency to engage in conspiratorial thinking. The four studies, including materials, hypotheses, and analyses, were pre-registered and all materials and data are accessible via the Open Science Framework (https://osf.io/8utsj/?view_only=364478f01fc548d0a42b9962918d871d). The differences between the four studies are summarized presently, and the methods are detailed below.

@@

In Study 1, we used questions drawn from questionnaires of scientific knowledge (e.g. "Are electrons smaller, larger, or the same size as atoms? [Smaller; Same size; Larger]"), supplemented by a ‘trick’ question ("Where do trees mainly draw the materials with which they create their mass? [Earth; Water; Air]"; correct answer: Air). In Studies 2 and 3, this last question was removed. The scientific facts used in Studies 1 to 3 represent long-established and basic knowledge. In Study 4 we used more recent much less basic scientific discoveries (e.g. "What is the electric charge of the Higgs Boson, as established in 2012? [1.602176634 × 10-19; 0; 3.2×10-19C]"; correct answer: 0, i.e. electrically neutral).

@@

In Study 1, we simply look at whether participants say that they accept the scientifically consensual answer. In the subsequent studies, we asked participants to explain why they disagreed with the scientific consensus. This revealed that a number of participants had made a mistake (misunderstanding, selecting the wrong answer). As a result, in Studies 3 and 4, participants who have indicated that they rejected the scientifically consensual answer were offered the option to revise their answer, or to keep rejecting it.

@@

The main goal of the present studies is descriptive: to find out whether participants who report not trusting science, or who believe in conspiracy theories, still accept most well-established scientific facts. However, we also tested two directional hypotheses (pre-registered as research questions in the Study 1):

@@

@@

           mean_knowledge_rate = paste0(mean_knowledge_rate*100, " %"))

@@

In all analyses reported in the main paper, we measure trust in science via a question selected from the Wellcome Global Monitor surveys [@wellcomeglobalmonitorWellcomeGlobalMonitor2018; @wellcomeglobalmonitorWellcomeGlobalMonitor2020]: "In general, would you say that you trust science a lot, some, not much, or not at all? [1 = Not at all, 2 = Not much, 3 = Some, 4 = A lot]". We chose this question as it seemed to be the most general one. In the ESM (sections \@ref(exp1) to \@ref(exp4)), we additionally report results for two alternative measures of trust included in our studies: Another from the WGM surveys ("How much do you trust scientists in this country? Do you trust them a lot, some, not much, or not at all? [1 = Not at all, 2 = Not much, 3 = Some, 4 = A lot]"), and one from the Pew Research Center [e.g. @funkKeyFindingsPublic2019] ("How much confidence do you have in scientists to act in the best interests of the public? [1-5; 1 = No confidence at all, 5 = A great deal of confidence]"), the latter having been used in a recent international study on trust in science [@colognaTrustScientistsTheir2024]. We selected these items so that we could compare the answers in our sample to global survey results. We find that all three items are generally highly correlated throughout all studies, and that our results reported here generally replicate when using either of the alternatives measures (with two exceptions: In Study 3, we find no correlation between acceptance and the Pew question; In Study 4, we find a correlation between knowledge and both alternative trust measures, but not with our main measure; see ESM, sections \@ref(exp3) and \@ref(exp4)).

@@

@@

Why did participants reject the scientific consensus? We collected a total of `r exp4_descriptives$justifications_n + exp3_descriptives$justifications_n + exp2_descriptives$justifications_n` answers (Studies 2: `r exp2_descriptives$justifications_n`; 3: `r exp3_descriptives$justifications_n`; 4: `r exp4_descriptives$justifications_n`) from `r exp4_descriptives$justifications_n_participants + exp3_descriptives$justifications_n_participants + exp2_descriptives$justifications_n_participants` (Studies 2: `r exp2_descriptives$justifications_n_participants`; 3: `r exp3_descriptives$justifications_n_participants`; 4: `r exp4_descriptives$justifications_n_participants`) participants to the open-ended questions on why they had rejected the scientific consensus on a particular question. Based on the answers, we created five categories (Table \@ref(tab:justifications)). All answers can be read in the ESM, sections \@ref(exp2) to \@ref(exp4).

@@

Why did participants say they accept the scientific consensus? In Studies 3 and 4--the vaccine hesitant samples--we had asked participants about cases in which they agreed with the scientific consensus. A total of `r exp3_descriptives$acceptance_n + exp4_descriptives$acceptance_n` (Studies 3: `r exp3_descriptives$acceptance_n`; 4: `r exp4_descriptives$acceptance_n`) participants answered this question. There were more participants saying they accepted the scientific consensus because they independently verified the fact (Studies 3: `r exp3_descriptives$acceptance_by_reason[["independent verification"]]$share`; 4: `r exp4_descriptives$acceptance_by_reason[["independent verification"]]$share` ), than participants saying it was because they trust scientists (Studies 3: `r exp3_descriptives$acceptance_by_reason[["trust in scientists"]]$share`; 4: `r exp4_descriptives$acceptance_by_reason[["trust in scientists"]]$share`)[^2]. Answers to a question about how they had done so can be found in the ESM, sections \@ref(exp3) and \@ref(exp4).

[^2]: `r exp3_descriptives$acceptance_by_reason[["other"]]$share` in Study 3 and `r exp4_descriptives$acceptance_by_reason[["other"]]$share` in Study 4 answered with "other" and gave an open-ended explanation (see ESM, sections \@ref(exp3) and \@ref(exp4)).

@@

In an exploratory analysis, we ran linear regressions to test whether there are differences between participants who said they had trusted science and those who said they had verified the information independently. Participants who said they accepted the consensus because of trust in scientists reported trusting science more (Studies 3: mean = `r exp3_descriptives$reason_means$'trust in scientists'$wgm_sciencegeneral_mean`; $\hat{\beta}_{\text{Trust}}$ = `r exp3_descriptives$reason_trust_model$'reason_agreementtrust in scientists'$estimate`, p `r exp3_descriptives$reason_trust_model$'reason_agreementtrust in scientists'$p.value` on a scale from 1 to 4; 4: mean = `r exp4_descriptives$reason_means$'trust in scientists'$wgm_sciencegeneral_mean`; $\hat{\beta}_{\text{Trust}}$ = `r exp4_descriptives$reason_trust_model$'reason_agreementtrust in scientists'$estimate`, p `r exp4_descriptives$reason_trust_model$'reason_agreementtrust in scientists'$p.value`) than those who said they verified independently (Studies 3 mean = `r exp3_descriptives$reason_means$'independent verification'$wgm_sciencegeneral_mean`; 4: mean = `r exp4_descriptives$reason_means$'independent verification'$wgm_sciencegeneral_mean`). We did not find a difference regarding acceptance (Studies 3: $\hat{\beta}_{\text{Acceptance}}$ = `r exp3_descriptives$reason_acceptance_model$'reason_agreementtrust in scientists'$estimate`, p `r exp3_descriptives$reason_acceptance_model$'reason_agreementtrust in scientists'$p.value` on a scale from 0 to 1; 4: $\hat{\beta}_{\text{Acceptance}}$ = `r exp4_descriptives$reason_acceptance_model$'reason_agreementtrust in scientists'$estimate`, p `r exp4_descriptives$reason_acceptance_model$'reason_agreementtrust in scientists'$p.value`) or regarding beliefs in conspiracy theories (Studies 3: $\hat{\beta}_{\text{BCTI}}$ = `r exp3_descriptives$reason_BCTI_model$'reason_agreementtrust in scientists'$estimate`, p `r exp3_descriptives$reason_BCTI_model$'reason_agreementtrust in scientists'$p.value` on a scale from 1 to 9; 4: $\hat{\beta}_{\text{BCTI}}$ = `r exp4_descriptives$reason_BCTI_model$'reason_agreementtrust in scientists'$estimate`, p `r exp4_descriptives$reason_BCTI_model$'reason_agreementtrust in scientists'$p.value`). We also did not find a difference in time spent on the survey in Study 3 ($\hat{\beta}_{\text{Time}}$ = `r exp3_descriptives$reason_time_model$'reason_agreementtrust in scientists'$estimate`, p `r exp3_descriptives$reason_time_model$'reason_agreementtrust in scientists'$p.value`; median = `r median(exp3_wide$duration_mins)` mins), but in Study 4, people who said they had accepted the consensus because they they trust scientists tended to spend on average two minutes less on the survey ($\hat{\beta}_{\text{Time}}$ = `r exp4_descriptives$reason_time_model$'reason_agreementtrust in scientists'$estimate`, p `r exp4_descriptives$reason_time_model$'reason_agreementtrust in scientists'$p.value`; median = `r median(exp4_wide$duration_mins)` mins).[^3] In Study 4, in which we used facts that participants were unlikely to have encountered before, we tracked whether people clicked on the source links we provided--a behavior that you would expect from people who report verifying facts independently. On average, participants clicked only on `r exp4_descriptives$reason_means$'trust in scientists'$sum_link_clicks_mean` links (out of 20 possible clicks) and there was no difference between the two groups ($\hat{\beta}_{\text{Clicks}}$ = `r exp4_descriptives$reason_clicks_model$'reason_agreementtrust in scientists'$estimate`, p `r exp4_descriptives$reason_clicks_model$'reason_agreementtrust in scientists'$p.value`).

[^3]: For these analyses, we excluded outliers who took over 30 mins for the survey, which was estimated to take around 10 mins, and for which the median time was seven minutes. As a result, we excluded one participant in Study 3 and four in Study 4. Significance levels are not affected by these exclusions.

@@

US participants were asked whether they accepted scientifically consensual answers on basic science questions.We found quasi-universal acceptance of basic science, with an overall rate of acceptance of `r combined_descriptives$mean_acceptance$Yes$share`, which remained very high for participants who declared not trusting science at all (`r combined_descriptives$acceptance_by_trust$'1'$Yes$share` of acceptance), or who endorsed theories blatantly violating scientific knowledge, such as flat earth (`r combined_descriptives$acceptance_by_conspiracy_belief$Yes$flatearth$'9'$share` of acceptance).

Overall, the present findings support the motivated rejection of science model [@lewandowskyMotivatedRejectionScience2016], in which “people tend to reject findings that threaten their core beliefs or worldview” (p. 217). A number of participants in our studies rejected basic tenets of science (on evolution, the shape of the earth, etc.). However, they still accepted the vast majority of basic scientific knowledge presented to them. This suggests that these participants had reasons to reject specific scientific knowledge, and that this rejection prompted them to express a lower trust in science when asked general questions on the topic. These participants did not appear to have general grounds for distrusting science, which should have led them to reject most or all of the science knowledge presented to them. These results are in line with several recent studies in which conspiracy theorists are not less susceptible to social influence in general [@pummererConspiracyBeliefsMajority; @altayConspiracyBelieversClaim2023].

Our findings are also relevant for the knowledge–attitudes model of trust in science [see, e.g., @bauerWhatCanWe2007]. The fact that even people with low science knowledge trust most of basic science might be taken as showing the power of even a modicum of science knowledge; on the other hand, it also means that attempts at increasing science knowledge in general might not have much effect on trust in science, by contrast with targeting the specific beliefs people are motivated to reject.

>>>>

---
title             : "Quasi-universal acceptance of basic science in the US"
shorttitle        : "Does anyone mistrust basic science?"
header-includes:  | # to prevent floats from moving past certain points (for the appendix)
  \usepackage{placeins}

author: 
  - name          : ""
    affiliation   : ""
affiliation:

  - id            : ""
    institution   : ""
abstract: |

  Substantial minorities of the population report a low degree of trust in science, or endorse conspiracy theories that violate basic scientific knowledge. This might indicate a wholesale rejection of science. In four studies, we asked 782 US participants questions about trust in science, conspiracy beliefs, and basic science (e.g. the relative size of electrons and atoms). Participants were provided with the scientifically consensual answer to the basic science questions, and asked whether they accept it. Acceptance of the scientific consensus was very high in the sample as a whole (95.1%), but also in every sub-sample (e.g., no trust in science: 87.3%; complete endorsement of flat Earth theory: 87.2%). This quasi-universal acceptance of basic science suggests that people are motivated to reject specific scientific beliefs, and not science as a whole. This could be leveraged in science communication.
keywords          : Trust in science; conspiracy theories; conspiracy thinking; science knowledge.
floatsintext      : yes
linenumbers       : no 
draft             : no
mask              : no
figurelist        : no
tablelist         : no
footnotelist      : no
classoption       : "doc" # "doc" for nice look, "man" for manuscripty
output: 
  papaja::apa6_docx: default
  papaja::apa6_pdf: 
    keep_tex: true
always_allow_html: true
appendix:
  - "appendix_acceptance_knowledge.Rmd"
  - "appendix_conspiracy_thinking.Rmd"
  - "appendix_trust.Rmd"
  - "appendix_conspiracy_belief.Rmd"
  - "appendix_knowledge.Rmd"
  - "appendix_by-question.Rmd"
  - "appendix_conspiracy-scales.Rmd"
  - "appendix_exp1.Rmd"
  - "appendix_exp2.Rmd"
  - "appendix_exp3.Rmd"
  - "appendix_exp4.Rmd"
bibliography: references.bib
---

@@

Although recent global evidence suggests that trust in science is moderately high [@colognaTrustScientistsTheir2024], it is far from being at ceiling. Large-scale polls have shown that people who report a high degree of trust in science are a minority in most countries, and they are outnumbered by people who have low trust in science in many areas, e.g., most of Africa and significant parts of Asia [@wellcomeglobalmonitorWellcomeGlobalMonitor2018; @wellcomeglobalmonitorWellcomeGlobalMonitor2020]. Moreover, trust in science has recently been declining in some countries [@alganTrustScientistsTimes2021; although see @wellcomeglobalmonitorPublicTrustScientists2021 and @funkPublicConfidenceScientists2020], including the US [@brianAmericansTrustScientists2023], where it is also increasingly polarizing [@gauchatPoliticizationSciencePublic2012; @krauseTrendsAmericansTrust2019a; @liPolarizationPublicTrust2022].

Besides low answers on general trust in science questions, another indicator of distrust in science is the belief in conspiracy theories that question the scientific consensus on issues such as vaccination, climate change, and even the shape of the Earth. Conspiracy theories–in the realm of science or elsewhere–typically accuse a small group of powerful people to pursue nefarious goals in secrecy [@douglasUnderstandingConspiracyTheories2019; @medeSciencerelatedPopulismConceptualizing2020]. Some of these conspiracy theories are widespread [@rutjensConspiracyBeliefsScience2022]. In 2023, a survey in eight different countries found that up to 24% of respondents agreed that “climate change is a hoax and scientists touting its existence are lying” [@stockemerUnderstandingClimateChange2024]. Similarly, in 2021, 40% of Americans believed that “the dangers of genetically-modified foods are being hidden from the public” [down from 45% in 2020, @uscinskiHaveBeliefsConspiracy2022]. People who hold such views not only reject the relevant scientifically consensual facts, but also tend to believe in other conspiracy theories [@lewandowskyRoleConspiracistIdeation2013; @hornseyPsychologicalRootsAntivaccination2018; @hornseyRelationshipsConspiratorialBeliefs2018] and tend to say that they distrust science more generally [@vranicDidMyOwn2022; @stockemerUnderstandingClimateChange2024]. Despite these correlations, the causal relationship between declaring general distrust in science and believing in one or several anti-science conspiracy theories is not clear. Although conspiracy thinking has been identified as a "root cause" of anti-science attitudes [@hornseyAttitudeRootsJiu2017] in the past, these claims rest largely on observational data.

What does this apparent lack of trust in science actually entail? Do people who say they do not trust science, or who believe in conspiracy theories at odds with well-established science, reject most of science? Or, on the contrary, do they object to a few specific facets of science, while still accepting the overwhelming majority of basic science?

A common conception of trust is a willingness to be vulnerable to another party, whether an individual, a group, or an institution [@mayerIntegrativeModelOrganizational1995a; @rousseauIntroductionSpecialTopic1998]. Accordingly, trust in science has been defined as “one’s willingness to rely on science and scientists (as representatives of the system) despite having a bounded understanding of science” [@wintterlinPredictingPublicTrust2022, p. 2]. Past research has disentangled this general concept of trust in science in various ways. Some research has identified different components of trust, the number of which varies, but which generally cover an epistemological and ethical dimension [@wilholtEpistemicTrustScience2013; @intemannScienceCommunicationPublic2023]. For example, @hendriksMeasuringLaypeoplesTrust2015 suggest distinguishing between expertise/competence, integrity, and benevolence, while @besleyReassessingVariablesUsed2021 add openness. Other research has highlighted differences in trust between scientific disciplines [@altenmullerExplainingPolarizedTrust2024; @gligoricHowSocialEvaluations2024; @gauchatCulturalCognitiveMappingScientific2018]. However, little research has assessed trust in specific scientific findings, besides contentious topics such as vaccines [e.g. @hornseyPsychologicalRootsAntivaccination2018], climate change [e.g. @stockemerUnderstandingClimateChange2024], evolution [e.g. @nadelsonTrustScienceScientists2015], genetically modified organisms (GMOs) [e.g. @fernbachExtremeOpponentsGenetically2019], or a combinations of such topics [see e.g @lewandowskyRoleConspiracistIdeation2013]. To the best of our knowledge, no research has investigated the extent to which people trust basic science facts (e.g. electrons are smaller than atoms). An extensive literature on science literacy has assessed whether people know such facts [@nationalacademiesofsciencesengineeringandmedicineScienceLiteracyConcepts2016], but not whether they accept the facts once presented to them.

Why does it matter whether trust in science in general is or is not related to trust in scientific findings? First, this question has theoretical implications. According to the most prominent explanation of (dis)trust in science–the deficit model–science knowledge is the main driver of attitudes towards science in general. A prediction of this model is that average trust in science on specific facts should be strongly associated with general trust in science. By contrast, a disconnect between the two would be in line with motivated reasoning accounts of trust in science. According to these accounts, science rejection serves to maintain coherence with other beliefs or behaviors [@lewandowskyMotivatedRejectionScience2016; @hornseyWhyFactsAre2020]. For example, someone might say they don't trust science in general because they are skeptical towards vaccines, not because they actually distrust most of science. A prediction of the motivated reasoning account is that general trust in science should be strongly correlated with conspiracy beliefs. Second, the present question has practical implications: Many communication attempts leverage the scientific consensus [e.g. on vaccination, climate change, etc., for review, see @vanstekelenburgScientificConsensusCommunicationContested2022; see also @veckalov27countryTestCommunicating2024]. These attempts are more likely to be successful if everyone trusts basic science than if some people reject science wholesale.

@@

In a series of four pre-registered online studies (total n = 782), we asked US participants questions about well-established, consensual scientific facts. For each question, we asked participants what they thought the correct answer was (testing their knowledge of science), we informed them of the scientifically accepted answer, and asked them whether they accepted it (measuring their trust in basic science). We also measured participants’ trust in science using standard measures, as well as their beliefs in various conspiracy theories and their tendency to engage in conspiratorial thinking. The four studies, including materials, hypotheses, and analyses, were pre-registered and all materials and data are accessible via the Open Science Framework (<https://osf.io/8utsj/?view_only=364478f01fc548d0a42b9962918d871d>). The differences between the four studies are summarized presently, and the methods are detailed below.

@@

In Study 1 (data collected on March 5, 2024), we used questions drawn from questionnaires of scientific knowledge (e.g. "Are electrons smaller, larger, or the same size as atoms? [Smaller; Same size; Larger]"), supplemented by a ‘trick’ question ("Where do trees mainly draw the materials with which they create their mass? [Earth; Water; Air]"; correct answer: Air). In Studies 2 (data collected on April 3, 2024) and 3 (data collected on April 22, 2024), this last question was removed. The scientific facts used in Studies 1 to 3 represent long-established and basic knowledge. In Study 4 (data collected on August 13, 2024) we used more recent, much less basic scientific discoveries (e.g. "What is the electric charge of the Higgs Boson, as established in 2012? [1.602176634 × 10-19; 0; 3.2×10-19C]"; correct answer: 0, i.e. electrically neutral).

@@

In Study 1, we simply looked at whether participants accept the scientifically consensual answer or not. In the subsequent studies, we asked participants to explain cases in which they disagreed with the scientific consensus. This revealed that a number of participants had made a mistake (misunderstanding, selecting the wrong answer). As a result, in Studies 3 and 4, participants who have indicated that they rejected the scientifically consensual answer were offered the option to revise their answer, or to keep rejecting it.

@@

The main goal of the present studies is descriptive: to find out whether participants who report not trusting science, or who believe in conspiracy theories, still accept most well-established scientific facts. However, based on our literature review, we also tested two directional hypotheses (pre-registered as research questions in the Study 1):

@@

# Compute the correlation between conspiracy belief and general trust
cor_trust_conspiracy_data <- combined_data |> 
  group_by(id) |> 
  summarise(BCTI_avg = mean(BCTI_avg, na.rm=TRUE), 
            wgm_sciencegeneral = mean(wgm_sciencegeneral, na.rm=TRUE))

cor_trust_conspiracy <- cor.test(cor_trust_conspiracy_data$wgm_sciencegeneral, cor_trust_conspiracy_data$BCTI_avg,
  use = "complete.obs") %>% text_ready()

@@

           mean_knowledge_rate = paste0(mean_knowledge_rate*100, " %")), 
  # overall correlation between conspiracy belief and general trust
  cor_trust_conspiracy = cor_trust_conspiracy 

@@

In all analyses reported in the main paper, we measure trust in science via a question selected from the Wellcome Global Monitor surveys [@wellcomeglobalmonitorWellcomeGlobalMonitor2018; @wellcomeglobalmonitorWellcomeGlobalMonitor2020]: "In general, would you say that you trust science a lot, some, not much, or not at all? [1 = Not at all, 2 = Not much, 3 = Some, 4 = A lot]". We chose this question as it seemed to be the most general one. In the ESM (sections \@ref(exp1) to \@ref(exp4)), we additionally report results for two alternative measures of trust included in our studies: Another from the WGM surveys ("How much do you trust scientists in this country? Do you trust them a lot, some, not much, or not at all? [1 = Not at all, 2 = Not much, 3 = Some, 4 = A lot]"), and one from the Pew Research Center [e.g. @funkKeyFindingsPublic2019] ("How much confidence do you have in scientists to act in the best interests of the public? [1-5; 1 = No confidence at all, 5 = A great deal of confidence]"), the latter having been used in a recent international study on trust in science [@colognaTrustScientistsTheir2024]. We selected these items so that we could compare the answers in our sample to global survey results. We find that all three items are highly correlated throughout all studies, and that our results reported here generally replicate when using either of the alternatives measures[^2] (see ESM).

[^2]: With two exceptions: In Study 3, we find no correlation between acceptance and the Pew question; In Study 4, we find a correlation between knowledge and both alternative trust measures, but not with our main measure; see ESM, sections \@ref(exp3) and \@ref(exp4)

@@

How do knowledge and acceptance relate to declared general trust in science and conspiracy belief (two strongly correlated variables, pooled r = `r combined_descriptives$cor_trust_conspiracy$estimate`, p `r combined_descriptives$cor_trust_conspiracy$p.value`)?

@@

n an exploratory analysis, we find that

While actual acceptance of scientific knowledge is only weakly correlated with trust in science,

Why did participants reject the scientific consensus? We collected a total of `r exp4_descriptives$justifications_n + exp3_descriptives$justifications_n + exp2_descriptives$justifications_n` answers (Studies 2: `r exp2_descriptives$justifications_n`; 3: `r exp3_descriptives$justifications_n`; 4: `r exp4_descriptives$justifications_n`) from `r exp4_descriptives$justifications_n_participants + exp3_descriptives$justifications_n_participants + exp2_descriptives$justifications_n_participants` (Studies 2: `r exp2_descriptives$justifications_n_participants`; 3: `r exp3_descriptives$justifications_n_participants`; 4: `r exp4_descriptives$justifications_n_participants`) participants to the open-ended questions on why they had rejected the scientific consensus on a particular question. Based on the answers, we created five categories (Table \@ref(tab:justifications)). All individual answers can be accessed in cleaned data sheets via the OSF project page.

@@

Why did participants say they accept the scientific consensus? In Studies 3 and 4--the vaccine hesitant samples--we had asked participants about cases in which they agreed with the scientific consensus. A total of `r exp3_descriptives$acceptance_n + exp4_descriptives$acceptance_n` (Studies 3: `r exp3_descriptives$acceptance_n`; 4: `r exp4_descriptives$acceptance_n`) participants answered this question. There were more participants saying they accepted the scientific consensus because they independently verified the fact (Studies 3: `r exp3_descriptives$acceptance_by_reason[["independent verification"]]$share`; 4: `r exp4_descriptives$acceptance_by_reason[["independent verification"]]$share` ), than participants saying it was because they trust scientists (Studies 3: `r exp3_descriptives$acceptance_by_reason[["trust in scientists"]]$share`; 4: `r exp4_descriptives$acceptance_by_reason[["trust in scientists"]]$share`)[^3]. Answers to a question about how they had done so can be found in the ESM, sections \@ref(exp3) and \@ref(exp4).

[^3]: `r exp3_descriptives$acceptance_by_reason[["other"]]$share` in Study 3 and `r exp4_descriptives$acceptance_by_reason[["other"]]$share` in Study 4 answered with "other" and gave an open-ended explanation.

@@

In an exploratory analysis, we ran linear regressions to test whether there are differences between participants who said they had trusted science and those who said they had verified the information independently. Participants who said they accepted the consensus because of trust in scientists reported trusting science more (Studies 3: mean = `r exp3_descriptives$reason_means$'trust in scientists'$wgm_sciencegeneral_mean`; $\hat{\beta}_{\text{Trust}}$ = `r exp3_descriptives$reason_trust_model$'reason_agreementtrust in scientists'$estimate`, p `r exp3_descriptives$reason_trust_model$'reason_agreementtrust in scientists'$p.value` on a scale from 1 to 4; 4: mean = `r exp4_descriptives$reason_means$'trust in scientists'$wgm_sciencegeneral_mean`; $\hat{\beta}_{\text{Trust}}$ = `r exp4_descriptives$reason_trust_model$'reason_agreementtrust in scientists'$estimate`, p `r exp4_descriptives$reason_trust_model$'reason_agreementtrust in scientists'$p.value`) than those who said they verified independently (Studies 3 mean = `r exp3_descriptives$reason_means$'independent verification'$wgm_sciencegeneral_mean`; 4: mean = `r exp4_descriptives$reason_means$'independent verification'$wgm_sciencegeneral_mean`). We did not find a difference regarding acceptance (Studies 3: $\hat{\beta}_{\text{Acceptance}}$ = `r exp3_descriptives$reason_acceptance_model$'reason_agreementtrust in scientists'$estimate`, p `r exp3_descriptives$reason_acceptance_model$'reason_agreementtrust in scientists'$p.value` on a scale from 0 to 1; 4: $\hat{\beta}_{\text{Acceptance}}$ = `r exp4_descriptives$reason_acceptance_model$'reason_agreementtrust in scientists'$estimate`, p `r exp4_descriptives$reason_acceptance_model$'reason_agreementtrust in scientists'$p.value`) or regarding beliefs in conspiracy theories (Studies 3: $\hat{\beta}_{\text{BCTI}}$ = `r exp3_descriptives$reason_BCTI_model$'reason_agreementtrust in scientists'$estimate`, p `r exp3_descriptives$reason_BCTI_model$'reason_agreementtrust in scientists'$p.value` on a scale from 1 to 9; 4: $\hat{\beta}_{\text{BCTI}}$ = `r exp4_descriptives$reason_BCTI_model$'reason_agreementtrust in scientists'$estimate`, p `r exp4_descriptives$reason_BCTI_model$'reason_agreementtrust in scientists'$p.value`). We also did not find a difference in time spent on the survey in Study 3 ($\hat{\beta}_{\text{Time}}$ = `r exp3_descriptives$reason_time_model$'reason_agreementtrust in scientists'$estimate`, p `r exp3_descriptives$reason_time_model$'reason_agreementtrust in scientists'$p.value`; median = `r median(exp3_wide$duration_mins)` mins), but in Study 4, people who said they had accepted the consensus because they they trust scientists tended to spend on average two minutes less on the survey ($\hat{\beta}_{\text{Time}}$ = `r exp4_descriptives$reason_time_model$'reason_agreementtrust in scientists'$estimate`, p `r exp4_descriptives$reason_time_model$'reason_agreementtrust in scientists'$p.value`; median = `r median(exp4_wide$duration_mins)` mins).[^4] In Study 4, in which we used facts that participants were unlikely to have encountered before, we tracked whether people clicked on the source links provided--a behavior that you would expect from people who report verifying facts independently. On average, participants clicked only on `r exp4_descriptives$reason_means$'trust in scientists'$sum_link_clicks_mean` links (out of 20 possible clicks) and there was no difference between the two groups ($\hat{\beta}_{\text{Clicks}}$ = `r exp4_descriptives$reason_clicks_model$'reason_agreementtrust in scientists'$estimate`, p `r exp4_descriptives$reason_clicks_model$'reason_agreementtrust in scientists'$p.value`).

[^4]: For these analyses, we excluded outliers who took over 30 mins for the survey, which was estimated to take around 10 mins, and for which the median time was seven minutes. As a result, we excluded one participant in Study 3 and four in Study 4. Significance levels are not affected by these exclusions.

@@

In four studies, we asked US participants whether they accepted scientifically consensual answers on basic science questions. We found quasi-universal acceptance of basic science, with an overall rate of acceptance of `r combined_descriptives$mean_acceptance$Yes$share`, which remained very high for participants who declared not trusting science at all (`r combined_descriptives$acceptance_by_trust$'1'$Yes$share` of acceptance), or who endorsed theories blatantly violating scientific knowledge, such as flat earth (`r combined_descriptives$acceptance_by_conspiracy_belief$Yes$flatearth$'9'$share` of acceptance).

This disconnect between declared general trust in science and average trust in (basic) scientific findings goes against predictions from the knowledge–attitudes model of trust in science [see, e.g., @bauerWhatCanWe2007], which posits that knowledge of scientific results is the main cause of trust in science. In line with past research [@allumScienceKnowledgeAttitudes2008], we find that both knowledge and acceptance of science are only weakly correlated with general trust in science. This supports contemporary definitions of science literacy, which suggest that for science literacy to be a meaningful concept for science attitudes, it needs to go beyond measuring knowledge of isolated science facts [@nationalacademiesofsciencesengineeringandmedicineScienceLiteracyConcepts2016].

Instead, our findings align with a motivated reasoning account of trust in science [@lewandowskyMotivatedRejectionScience2016], in which “people tend to reject findings that threaten their core beliefs or worldview” (p. 217). A number of participants in our studies endorsed specific conspiracy theories questioning basic tenets of science (on evolution, the shape of the earth, etc.), while still accepting the vast majority of basic scientific facts presented to them. This suggests that these participants had reasons to reject only specific scientific knowledge, and that this rejection prompted them to express a lower trust in science when asked general questions on the topic. Consistent with this explanation, we found a strong association between belief in the anti-science conspiracy theories and general trust in science.
